\chapter{Collisions Between Particles}

% ============ Problem 1 ============ %

\begin{problem}
{
Find the relation between the angles $\theta_1$, $\theta_2$ (in the $L$ system) after disintegrating into two particles.
}
{
The angles of the particles, in the $C$ system, are related by
\begin{equation*}
\theta_0 = \theta_{10} = \pi - \theta_{20},
\end{equation*}
where $\theta_0$ as been defined to simplify the notation. From the formulae (16.5) of the book, we have
\begin{align*}
    V + v_{10}\cos{\theta_0} &= v_{10} \sin{\theta_0} \cot{\theta_1} \\
    V - v_{20}\cos{\theta_0} &= v_{20} \sin{\theta_0} \cot{\theta_2} .
\end{align*}
We must eliminate $\theta_0$ from the two equations above. To do so, we can first solve for $\cos{\theta_0}$ and $\sin{\theta_0}$ which give us
\begin{align}
    &\sin{\theta_0} = \frac{V + v_{10}\cos{\theta_0}}{v_{10}\cot{\theta_1}} = \frac{V - v_{20}\cos{\theta_0}}{v_{20}\cot{\theta_2}} \label{C4P1_sin} \\
    \Rightarrow& Vv_{20}\cot{\theta_2} + v_{10}v_{20}\cos{\theta_0}\cot{\theta_2} = Vv_{10}\cot{\theta_1} - v_{10}v_{20}\cos{\theta_0}\cot{\theta_1} \nonumber \\
    \Rightarrow& \cos{\theta_0} = \frac{V\left( v_{10}\cot{\theta_1} - v_{20}\cot{\theta_2} \right)}{v_{10}v_{20} \left( \cot{\theta_1} + \cot{\theta_2} \right)}.  \label{C4P1_cos}
\end{align}
Using the sum of square of \eqref{C4P1_sin} and \eqref{C4P1_cos}, we get
\begin{align*}
    &\left( \frac{V + v_{10}\cos{\theta_0}}{v_{10}\cot{\theta_1}} \right)^2 + \left( \frac{V\left( v_{10}\cot{\theta_1} - v_{20}\cot{\theta_2} \right)}{v_{10}v_{20} \left( \cot{\theta_1} + \cot{\theta_2} \right)} \right)^2 = 1 \\
    \Rightarrow& \left( v_{10} + v_{20} \right)^2 + \left( v_{10}\cot{\theta_1} - v_{20}\cot{\theta_2} \right)^2 = \frac{v_{10}^2v_{20}^2}{V^2} \left( \cot{\theta_1} + \cot{\theta_2} \right)^2 \\
    \Rightarrow& v_{10}^2 \csc{\theta_1}^2 + v_{20}^2 \csc{\theta_2}^2 - 2 v_{10}v_{20}\cot{\theta_1}\cot{\theta_2} = \frac{v_{10}^2v_{20}^2\sin^2{(\theta_1+\theta_2)}}{V^2\sin^2{\theta_1}\sin^2{\theta_2}} \\ 
    \Rightarrow& v_{20}^2 \sin^2{\theta_1} + v_{10}^2 \sin^2{\theta_2} - 2 v_{10}v_{20}\sin{\theta_1}\sin{\theta_2}\cos{\theta_1}\cos{\theta_2} = \frac{v_{10}^2v_{20}^2}{V^2}\sin^2{(\theta_1+\theta_2)}.
\end{align*}
Using the equation (16.2) from the book and the relation $v_{10}/v_{20} = m_2/m_1$, we obtain
}
{
\begin{equation*}
    \frac{m_2}{m_1} \sin^2{\theta_2} + \frac{m_1}{m_2} \sin^2{\theta_1} - 2 \sin{\theta_1}\sin{\theta_2}\cos{\theta_1}\cos{\theta_2} = \frac{2\epsilon}{(m_1+m_2)V^2} \sin^2{(\theta_1+\theta_2)}.
\end{equation*}
}
\end{problem}

% ============ Problem 2 ============ %

\begin{problem}
{
    Find the angular distribution of the resulting particles in the $L$ system.
}
{
The distribution of the resulting particles with respect to the angle $\theta_0$ is given by the equation (16.7) of the book, \ie
\begin{equation*}
    \frac{1}{2}\sin{\theta_0}d\theta_0 = \frac{1}{2}\mathrm{d}(\cos{\theta_0}).
\end{equation*}
Using equation (16.6) of the book yield
\begin{equation*}
    \frac{1}{2}\mathrm{d}(\cos{\theta_0}) = -\frac{V}{v_0}\sin{\theta}\cos{\theta}\mathrm{d}\theta \pm \left( \sin{\theta}\sqrt{1-(V^2/v_0^2)\sin^2{\theta}} + \cos{\theta}\frac{(V^2/v_0^2)\sin{\theta}\cos{\theta}}{\sqrt{1-(V^2/v_0^2)\sin^2{\theta}}} \right) \mathrm{d}\theta.
\end{equation*}
After simplification, the angular distribution become
}
{
\begin{equation*}
    \frac{1}{2}\sin{\theta} \mathrm{d}\theta \left( 2\frac{V}{v_0}\cos{\theta} \pm \frac{1+(V^2/v_0^2)\cos{2\theta}}{\sqrt{1-(V^2/v_0^2)\sin^2{\theta}}} \right)
\end{equation*}
}
\end{problem}

% ============ Problem 3 ============ %
% See https://www.physicsforums.com/threads/trigonometric-function-mechanics-landau.962697/

\begin{problem}
{
Determine the range of possible values of the angle $\theta$ between the directions of motion of the two resulting particles in the $L$ system.
}
{
The tangent of the angle $\theta = \theta_1 + \theta_2$ is
\begin{equation*}
    \tan{\theta} = \frac{\tan{\theta_1} + \tan{\theta_2}}{1 - \tan{\theta_1}\tan{\theta_2}}.
\end{equation*}
Using equation (16.5) of the book to get $\tan{\theta_1}$ and $\tan{\theta_1}$, and using $\theta_0 = \theta_{10} = \pi - \theta_{20}$, the last equation become
\begin{equation*}
    \tan{\theta} = \frac{(v_{10}+v_{20})V\sin{\theta_0}}{V^2+(v_{10}-v_{20})V\cos{\theta_0} - v_{10}v_{20}}
\end{equation*}
or equivalently
\begin{equation*}
    \cot{\theta} = \frac{V^2+(v_{10}-v_{20})V\cos{\theta_0} - v_{10}v_{20}}{(v_{10}+v_{20})V\sin{\theta_0}}.
\end{equation*}
In order to find the extrema, we take the derivative with respect to $\theta_0$,
\begin{align*}
    &\pd{\cot{\theta}}{\theta_0} = 0 = \frac{(v_{20}-v_{10})V^2 - V\cos{\theta_0}(V^2-v_{10}v_{20})}{(v_{10}+v_{20})V^2\sin^2{\theta_0}}\\
    \Rightarrow &\cos{\theta_0} = \frac{(v_{20}-v_{10})V}{V^2-v_{10}v_{20}}.
\end{align*}
We know that $-1 \leq \cos{\theta_0} \leq 1$, thus $v_{10} < V < v_{20}$. Moreover, we could calculate the derivative of the cotangent with respect to $ V $ and find the expression:
\begin{equation*}
    \frac{\partial \cot{\theta}}{\partial V} = \frac{V^2 + v_{10}v_{20}}{(v_{10}+v_{20})V^2 \sin{\theta_0}}
\end{equation*}
which is positive for all values of $ V $.

This is why $ \cot{\theta} $ may assume all the values between 0 and $ \pi $ when $ v_{10} < V < v_{20} $, because $ \frac{\partial \cot{\theta}}{\partial \theta_0} \neq 0 $ and $ \frac{\partial \cot{\theta}}{\partial V} \neq 0 $ for all $ V, \theta_0 $ in that interval. It is easy to see that if $ V $ is kept constant and $ \theta_0 $ is varied between 0 and $ \pi $, $ \cot{\theta} $ spans the interval $ (-\infty, +\infty) $; similarly, if $ \theta_0 $ is kept constant and $ V $ is varied between 0 and $ +\infty $, $ \cot{\theta} $ spans the interval $ (-\infty, +\infty) $.

When $ V \leq v_{10} $, it is easy to see that $ \cot{\theta} $ has a maximum value. In fact,
\begin{equation*}
    \lim_{\theta_0 \to 0^+} \cot(x) = -\infty, \quad \lim_{\theta_0 \to \pi^-} \cot(x) = -\infty
\end{equation*}
and the maximum value of the cotangent is reached when:
\begin{equation*}
    \cos{\theta_0} = \frac{(v_{20} - v_{10})V}{V^2 - v_{10}v_{20}}
\end{equation*}
The calculation of the maximum value leads to:
\begin{equation*}
    \cot{\theta} = \frac{V^2 + (v_{10} - v_{20})V\cos{\theta_0} - v_{10}v_{20}}{(v_{10}+v_{20})V \sin{\theta_0}} = -\frac{\sqrt{(V^2 - v_{10}v_{20})^2 - (v_{20} - v_{10})^2 V^2}}{(v_{10} + v_{20})V}
\end{equation*}
Then, after transforming the cotangent into a sine, one gets:
\begin{equation*}
    \sin{\theta} = \frac{(v_{20} + v_{10})V}{V^2 + v_{10}v_{20}}
\end{equation*}
If we put:
\begin{equation*}
    \alpha = \sin^{-1}\left( \frac{(v_{20} + v_{10})V}{V^2 + v_{10}v_{20}} \right)
\end{equation*}
then, in the interval where $ V \leq v_{10} $, we have the following ranges for $ \theta $: $ \pi - \alpha \leq \theta \leq \pi $.
Similarly, when $ V \geq v_{20} $, it is easy to see that $ \cot{\theta} $ has a minimum value. In fact,
\begin{equation*}
    \lim_{\theta_0 \to 0^+} \cot(x) = +\infty, \quad \lim_{\theta_0 \to \pi^-} \cot(x) = +\infty
\end{equation*}
and the minimum value of the cotangent is reached when $ \theta = \alpha $. In this case, the ranges for $ \theta $ are $ 0 \leq \theta \leq \alpha $ (the equality $ \theta = 0 $ is satisfied only when $ V \to +\infty $). The final solution is thus
}
{
\begin{align*}
    &\text{if } v_{10} < V < v_{20} \quad \Rightarrow \quad 0 < \theta < \pi \\
    &\text{if } V \leq v_{10} \quad \Rightarrow \quad \pi - \alpha \leq \theta \leq \pi \\
    &\text{if } V \geq v_{20} \quad \Rightarrow \quad 0 \leq \theta \leq \alpha
\end{align*}
where
\begin{equation*}
    \alpha = \sin^{-1} \left( \frac{(v_{20} + v_{10})V}{V^2 + v_{10}v_{20}} \right).
\end{equation*}
}
\end{problem}

% ============ Problem 4 ============ %

\begin{problem}
{
Express the velocity of each particle after a collision between a moving particle ($m_1$) and another at rest ($m_2$) in terms of their directions of motion in the $L$ system.
}
{
To solve this problem with need to use the Fig. 16 of the book. First, using the law of sines, the magnitude of the momentum of the second particle after the collision is
\begin{equation*}
p_2' = \mathrm{OB}\frac{\sin{\chi}}{\sin{\theta_2}} = mv\frac{\sin{\chi}}{\cos{(\chi/2)}} = 2mv\sin{\frac{\chi}{2}}.
\end{equation*}
We get the equation (17.5) of the book if we isolate the velocity
\begin{equation*}
    v_2' = \frac{2m_1v}{m_1+m_2}\sin{\frac{\chi}{2}}.
\end{equation*}
Second, using the law of cosines, we get
\begin{align*}
    &\mathrm{OC}^2 = \mathrm{AO}^2 + p_1'^2 - 2\mathrm{AO}p_1'\cos{\theta_1} \\
    \Rightarrow &p_1' = \mathrm{AO}\cos{\theta_1} \pm \sqrt{\mathrm{AO}^2\cos^2{\theta_1} + \mathrm{OC}^2 - \mathrm{AO}^2} \\
    \Rightarrow &v_1' = \frac{m_1\cos{\theta_1} \pm \sqrt{m_2^2-m_1^2\sin^2{\theta_1}}}{m_1+m_2}v_1.
\end{align*}
For the case where $m_2 > m_1$ we also get the equation (17.5) of the book, but when $m_1 > m_2$, the radical may have either sign. To summarize, the answer is
}
{
\begin{align*}
    &v_2' = \frac{2m_1v}{m_1+m_2}\sin{\frac{\chi}{2}} \\
    &v_1' = \frac{m_1\cos{\theta_1} \pm \sqrt{m_2^2-m_1^2\sin^2{\theta_1}}}{m_1+m_2}v_1.
\end{align*}
}
\end{problem}

% ============ Problem 5 ============ %
% Need to make a figure

\begin{problem}
{
Determine the effective cross-section for scattering of particles from a perfectly rigid sphere of radius $a$ (\ie when the interaction is such that $U=\infty$ for $r<a$ and $U=0$ for $r>a$).
}
{
From Fig. 19 of the book, we can see
\begin{equation}
    \rho = a\sin{\phi_0}=a\sin{\frac{1}{2}(\pi-\chi)}=a\cos{\frac{1}{2}\chi}. \label{C4P5_rho}
\end{equation} 
Subsituing \eqref{C4P5_rho} in equation (18.7) or (18.8) of the book, we have
\begin{equation}
    \mathrm{d}\sigma=\frac{1}{2}\pi a^2 \sin{\chi}\mathrm{d}\chi=\frac{1}{4}a^2\mathrm{d}o. \label{C4P5_disgma}
\end{equation}
If we integrate $\mathrm{d}\sigma$ over all angles, we find
\begin{equation*}
    \sigma = \frac{1}{2}\pi a^2  \int_{-\pi/2}^{\pi/2} \sin{\chi}\mathrm{d}\chi =\pi a^2, 
\end{equation*}
which is the projection of the sphere perpendicular to the the beam (\ie the impact area). We need to express \eqref{C4P5_disgma} in the $L$ system using the equation (17.4) of the book. If this equation is solve for $\cos{\theta_1}$, we get
\begin{align*}
    &\tan{\theta_1} = \frac{m_2\sin{\chi}}{m_1+m_2\cos{\chi}}\\
    &\Rightarrow \tan^2{\theta_1}\left( \frac{m_1}{m_2} + \cos{\chi} \right)^2 = 1 - \cos^2{\chi} \\
    &\Rightarrow \cos{\chi} = -\frac{m_1}{m_2}\sin^2{\theta_1} \pm \cos{\theta_1}\sqrt{1-\frac{m_1^2}{m_2^2}\sin^2{\theta_1}} \\
    &\Rightarrow \frac{1}{2}\sin{\chi}\mathrm{d}\chi = \frac{1}{2}\sin{\theta_1} \mathrm{d}\theta_1 \left( 2\frac{m_1}{m_2}\cos{\theta_1} \pm \frac{1+(m_1^2/m_2^2)\cos{2\theta_1}}{\sqrt{1-(m_1^2/m_2^2)\sin^2{\theta_1}}} \right).
\end{align*}
We thus get
}
{
\begin{align*}
    \mathrm{d}\sigma_1 &= \frac{1}{4}a^2\left( 2\frac{m_1}{m_2}\cos{\theta_1} \pm \frac{1+(m_1^2/m_2^2)\cos{2\theta_1}}{\sqrt{1-(m_1^2/m_2^2)\sin^2{\theta_1}}} \right) \mathrm{d}o_1 \\
    \mathrm{d}\sigma_2 &= a^2|\cos{\theta_2}|\mathrm{d}o_2.
\end{align*}
}
\end{problem}

% ============ Problem 6 ============ %

\begin{problem}
{
Express the effective cross-section (see last problem) as a function of the energy $\epsilon$ lost by a scattered particle.
}
{
The energy loss by a particle of mass $m_1$ is equal to the energy gained by the sphere of mass $m_2$. Knowing that the sphere is initially at rest, the energy gained is
\begin{equation*}
    \epsilon = E_2' = \frac{1}{2} m_2 v_2'^2 = \frac{2 m_1^2 m_2}{\left( m_1+m_2 \right)^2} v_\infty^2\sin{\frac{1}{2}\chi} = \epsilon_{max} \sin{\frac{1}{2}\chi}.
\end{equation*}
Using the double angle identity, the variation of $\epsilon$ is
\begin{equation*}
    \mathrm{d}\epsilon = \epsilon_{max}\sin{\frac{1}{2}\chi}\cos{\frac{1}{2}\chi}\mathrm{d}\chi = \frac{1}{2}\epsilon_{max}\sin{\chi}\mathrm{d}\chi.
\end{equation*}
Using \eqref{C4P5_disgma} from the last problem, the effective cross section can thus be express as
}
{
\begin{equation*}
    \mathrm{d}\sigma = \frac{\pi a^2 \mathrm{d}\epsilon}{\epsilon_{max}}
\end{equation*}
}
\end{problem}

% ============ Problem 7 ============ %

\begin{problem}
{
Find the effective cross-section as a function of the velocity $v_\infty$ for particles scattered in a field $U \sim r^{-n}$.
}
{
For a central potential field, the scattering angle is determined by the integral
\begin{equation*}
    \chi = \pi - 2\int_{r_{min}}^\infty \frac{\rho \mathrm{d}r}{r^2 \sqrt{1 - \frac{\rho^2}{r^2} - \frac{2U(r)}{mv_\infty^2}}}.
\end{equation*}
Subsituing $U(r) = \alpha r^{-n}$ and define the dimensionless variable $x = \frac{r}{\rho}$, then
\begin{equation*}
    \chi = \pi - 2\int_{x_{min}}^\infty \frac{\mathrm{d}x}{x^2 \sqrt{1 - \frac{1}{x^2} - \frac{2\alpha}{mv_\infty^2\rho^nx^n}}}.
\end{equation*}
Notice the dimensionless combination $\xi = \frac{2\alpha}{mv_\infty^2\rho^n}$, the integral simplifies to
\begin{equation*}
    \chi = \pi - 2\int_{x_{min}}^\infty \frac{\mathrm{d}x}{x^2 \sqrt{1 - \frac{1}{x^2} - \frac{\xi}{x^n}}}.
\end{equation*}
Thus, the scattering angle depends only on $\xi$,
\begin{equation*}
    \chi = f(\xi),
\end{equation*}
where $f(\xi)$ is a dimensionless function determined by the integral. To invert this relationship, solve for $\rho$ in terms of $\chi$ and $v_\infty$,
\begin{equation*}
    \xi = \frac{2\alpha}{mv_\infty^2\rho^n} \Rightarrow \rho = \left( \frac{2\alpha}{mv_\infty^2\xi} \right)^{1/n}.
\end{equation*}
Let $\xi = g(\chi)$, where $g(\chi)$ is a function encoding the angular dependence. Then,
\begin{align*}
    \rho &= \left( \frac{2\alpha}{mv_\infty^2} \right)^{1/n} \left( g(\chi) \right)^{-1/n} \\
    \Rightarrow \rho &\sim v_\infty^{-2/n}\left( g(\chi) \right)^{-1/n}.
\end{align*}
The differential cross-section is thus
\begin{align*}
    \mathrm{d}\sigma &= \frac{\rho}{\sin{\chi}} \left| \frac{\mathrm{d}\rho}{\mathrm{d}\chi} \right| \mathrm{d}o \\
    \mathrm{d}\sigma &\sim v_\infty^{-4/n} \frac{F(\chi)}{\sin{\chi}} \mathrm{d}o,
\end{align*}
where $F(\chi) = \left| \frac{\mathrm{d}}{\mathrm{d}\chi} \left( g(\chi) \right)^{-1/n} \right|$. Integrate over the solid angle
\begin{equation*}
    \sigma \sim v_\infty^{-4/n} \int F(\chi) \mathrm{d}\chi.
\end{equation*}
Knowing that the integral depends only on $n$, not $v_\infty$, we have
}
{
\begin{equation*}
    \sigma \sim v_\infty^{-4/n}
\end{equation*}
}
\end{problem}

% ============ Problem 8 ============ %

\begin{problem}
{
Determine the effective cross-section for a particle to ``fall" to the centre of a field $U=-\alpha/r^2$.
}
{
}
{
}
\end{problem}